---
title: "Untitled"
output: html_document
---

# 10.5 Exercises with Tibbles

1. How can you tell if an object is a tibble? (Hint: try printing mtcars, which is a regular data frame).

A tibble shows the types of the columns.

```{r}
as_tibble(mtcars)
```


2. Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?

With a dataframe:
```{r}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

df$x -- This does partial matching of x to xyz (undesired feature). Moreover, when creating the df, the variable xyz was automatically turned into a factor.

df[, "xyz"] -- this prints all rows of column named xyz. Somehow more complicated (is it?) of something that could be done with df[["xyz"]] with a tibble.

df[, c("abc", "xyz")] -- same comment as before.

If we were using a tibble, instead:
```{r}
tb <- tibble(abc = 1, xyz = "a")
tb$x # Gives an error: Unknown column 'x'
tb[, "xyz"] # xyz is simply a char variable and not a factor.
tb[, c("abc", "xyz")] # Same comment as above.
```

3. If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the reference variable from a tibble?

Either by var$mpg, var[["mpg"]] or var[[ncol(var)]].


4. Practice referring to non-syntactic names in the following data frame by:

Extracting the variable called 1.

Plotting a scatterplot of 1 vs 2.

Creating a new column called 3 which is 2 divided by 1.

Renaming the columns to one, two and three.

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

# Extracting the variable called 1
annoying$`1` 

# Plotting a scatterplot of 1 vs 2.
plot(annoying$`1`, annoying$`2`, main="Scatterplot", 
  	xlab="1", ylab="2", pch=19)

# Creating a new column called 3 which is 2 divided by 1.
annoying$`3` <- annoying$`2` / annoying$`1`
annoying

# Renaming the columns to one, two and three.
names(annoying) <- c("one","two","three")
annoying
```


5. What does tibble::enframe() do? When might you use it?

NOT SURE!!
From help: A helper function that converts named atomic vectors or lists to two-column data frames. For unnamed vectors, the natural sequence is used as name column.


6. What option controls how many additional column names are printed at the footer of a tibble?

NOT SURE!!

options(tibble.width = Inf) will always print all columns, regardless of the width of the screen.

```{r}
tb_flights <- as_tibble(nycflights13::flights)
options(tibble.max_extra_cols = 5)
tb_flights <- as_tibble(nycflights13::flights)
"print"(tb_flights, width = 3)
```

