---
title: "Untitled"
output: html_document
---

# 13.3.1 Exercises -- keys

1. Add a surrogate key to flights.

```{r}
library(tidyverse)
library(nycflights13)
```

```{r}
(flights1 <- mutate(flights, key = row_number()))
```


2. Identify the keys in the following datasets

Lahman::Batting,
```{r}
if("Lahman" %in% rownames(installed.packages()) == FALSE) {install.packages("Lahman")}
Lahman::Batting

Lahman::Batting %>% 
  count(playerID, yearID, stint) %>% 
  filter(n > 1)
```


babynames::babynames
```{r}
if("babynames" %in% rownames(installed.packages()) == FALSE) {install.packages("babynames")}

babynames::babynames %>%
  count(name,year) %>%
  filter(n > 1)
```


nasaweather::atmos
```{r}
if("nasaweather" %in% rownames(installed.packages()) == FALSE) {install.packages("nasaweather")}

nasaweather::atmos

nasaweather::atmos %>%
  count(lat,long,year,month) %>%
  filter(n > 1)
```


fueleconomy::vehicles
```{r}
if("fueleconomy" %in% rownames(installed.packages()) == FALSE) {install.packages("fueleconomy")}

fueleconomy::vehicles 
fueleconomy::vehicles %>%
  count(id) %>%
  filter(n > 1)
```

ggplot2::diamonds
(You might need to install some packages and read some documentation.)
```{r}
if("ggplot2" %in% rownames(installed.packages()) == FALSE) {install.packages("ggplot2")}

ggplot2::diamonds
ggplot2::diamonds %>%
  count(carat,cut,color,clarity,depth,table,price,x,y,z) %>%
  filter(n > 1)

# There are no keys. Even using all the columns, there are 2 of each.
```


3. Draw a diagram illustrating the connections between the Batting, Master, and Salaries tables in the Lahman package. Draw another diagram that shows the relationship between Master, Managers, AwardsManagers.

```{r}
if("Lahman" %in% rownames(installed.packages()) == FALSE) {install.packages("Lahman")}

Lahman::Batting 
# Key: playerID, yearID, stint

Lahman::Master
# Cannot see it here but connects to Batting through playerID>
Lahman::Salaries 
# Connects to Batting through PlayerID, yearID, lgID, teamID.
```

4. How would you characterise the relationship between the Batting, Pitching, and Fielding tables?

```{r}
Lahman::Pitching 
# One player can do many pitching. Each pitching is done by 1 player.
Lahman::Fielding
# One player can do many fielding. Each fielding is done by 1 player.
```


# 13.4.6 Exercises -- Mutating joins

1. Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:
(Don’t worry if you don’t understand what semi_join() does — you’ll learn about it next.)
```{r}
airports %>%
  semi_join(flights, c("faa" = "dest")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

You might want to use the size or colour of the points to display the average delay for each airport.

```{r}
if("maps" %in% rownames(installed.packages()) == FALSE) {install.packages("maps")}

# Compute the average delay by destination
(flights %>%
  filter(arr_delay != "NA", dep_delay !="NA") %>%
  group_by(dest) %>%
  summarise(avg_delay = mean(c(arr_delay,dep_delay))) %>%
  left_join(airports, c("dest" = "faa")))


```

Add the location of the origin and destination (i.e. the lat and lon) to flights.

Is there a relationship between the age of a plane and its delays?

What weather conditions make it more likely to see a delay?

What happened on June 13 2013? Display the spatial pattern of delays, and then use Google to cross-reference with the weather.
